
---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---



# Reproducible Research Peer Assessment 1

## Loading and preprocessing the data



```{r}
options(scipen = 4, digits = 2)

setwd("C:/Users/JWilkinson/represearch/RepData_PeerAssessment1")
df <- read.csv(unz("activity.zip", "activity.csv"))
df$date<- as.Date(df$date, format = "%Y-%d-%m")
```

## What is mean total number of steps taken per day?

```{r}

df2 <- aggregate(steps ~ date, data = df, sum, na.rm = TRUE)
hist(df2$steps, xlab = "Steps", main = "Steps per day", col = "peachpuff", breaks = 9)
smean <- mean(df2$steps)
smedian<- median(df2$steps)
abline(v = smean, col = "blue")
abline(v = smedian, col = "red")
legend(x = "topright",c("Mean", "Median"), col = c("blue", "red"), lwd = c(1,1))

```


As can be seen in the histogram above the mean steps per day is `r smean`, and the median is `r smedian`.

## What is the average daily activity pattern?

```{r}

agdf <- aggregate(df, by = list(df$interval), FUN = mean, na.rm = TRUE)
plot(agdf$steps~agdf$interval, type = "l", main = "Steps per interval", xlab = "Interval", ylab = "Steps")
wm <- which.max(agdf[,2])
maxint <- agdf[wm,"interval"]
maxstep <- agdf[wm,"steps"]
abline(v = maxint, col = "orange", lwd = 2)

```



As noted in the graph above, interval `r maxint` has the highest average steps at `r maxstep`.

## Imputing missing values

```{r}

keep <- !is.na(df$date)
df3 <- df[keep,]
missingsteps <- nrow(df3) - nrow(df2)
totmissing <- nrow(df) - nrow(df2)

#due to the uniform nature of the spread of NA values a simple arithmetic mean by interval for NAs replacement was used.

library(data.table)
filldf <- data.table(agdf, key = "interval")
nadf <- data.table(df3, key = "interval")
nadf$steps <- as.numeric(nadf$steps)
outdf <- nadf[filldf]

#replacing na values with means by interval
outdf[is.na(steps), steps:= steps.1]

df4 <- aggregate(steps ~ date, data = outdf, sum)

smean2 <- mean(df4$steps)
smedian2 <- median(df4$steps)
hist(df4$steps, xlab = "Steps", main = "Steps per day", col = "lightgreen", breaks = 9)
abline(v = smean2, col = "blue",lwd = 2, lty = 2)
abline(v = smedian2, col = "red", lwd = 2)
legend(x = "topright",c("Mean", "Median"), col = c("blue", "red"), lwd = c(2,2), lty = c(2,1))

```



The total rows in our original dataset with NA values is `r totmissing`.  The number of rows of data where dates were recorded but there are NA's for steps values is `r missingsteps`.  It is these rows with dates recorded that will be used going forward with mean for each interval replacing NA values.  The uniform spread of NA's across intervals should minimize the impact of replacement in this manner.  As can be seen in the histogram the new mean is `r smean2` which is nearly the same (less than .1% difference) and the median of `r smedian2` is altered by about 1%.


## Are there differences in activity patterns between weekdays and weekends?


```{r}

library("plyr")
library("lattice")

outdf$day <- weekdays(outdf$date)
outdf$wkend <- ifelse(outdf$day %in% c("Saturday", "Sunday"),"weekend", "weekday")
avgmeandf <- ddply(outdf, c("interval","wkend"), summarise, mean = mean(steps))
xyplot(mean ~ interval | wkend, avgmeandf, type = "l", layout = c(1,2), ylab = "Average Steps", xlab = "Interval")

```

Average steps per day for weekends vs weekdays over 5 minute intervals outlined above.
