## Reproducible Research Peer Assessment 1

```{r}

setwd("C:/Users/JWilkinson/represearch/RepData_PeerAssessment1")
df <- read.csv(unz("activity.zip", "activity.csv"))
df$date<- as.Date(df$date, format = "%Y-%d-%m")

df2 <- na.omit(df)
hist(df2$steps, xlab = "Steps", main = "Steps per day", col = "peachpuff")
smean <- mean(df2$steps)
smedian<- median(df2$steps)
abline(v = smean, col = "blue")
abline(v = smedian, col = "red")
legend(x = "topright",c("Mean", "Median"), col = c("blue", "red"), lwd = c(1,1))

```

As can be seen in the histogram above the mean steps per day is `r smean`, and the median is `r smedian`.



```{r}

agdf <- aggregate(df, by = list(df$interval), FUN = mean, na.rm = TRUE)
plot(agdf$steps~agdf$interval, type = "l", main = "Steps per interval", xlab = "Interval", ylab = "Steps")
wm <- which.max(agdf[,2])
maxint <- agdf[wm,"interval"]
maxstep <- agdf[wm,"steps"]
abline(v = maxint, col = "orange")


```

As noted in the graph above, interval `r maxint` has the highest average steps at `r maxstep`.



```{r}

keep <- !is.na(df$date)
df3 <- df[keep,]
missingsteps <- nrow(df3) - nrow(df2)
totmissing <- nrow(df) - nrow(df2)

#due to the uniform nature of the spread of NA values a simple arithmetic mean by interval for NAs replacement was used.

library(data.table)
filldf <- data.table(agdf, key = "interval")
nadf <- data.table(df3, key = "interval")
nadf$steps <- as.numeric(nadf$steps)
outdf <- nadf[filldf]
outdf[is.na(steps), steps:= steps.1]
smean2 <- mean(outdf$steps)
smedian2 <- median(outdf$steps)
hist(outdf$steps, xlab = "Steps", main = "Steps per day", col = "lightgreen")
abline(v = smean2, col = "blue",lwd = 2)
abline(v = smedian2, col = "red", lwd = 2)
legend(x = "topright",c("Mean", "Median"), col = c("blue", "red"), lwd = c(2,2))

```

The total rows in our original dataset with NA values is `r totmissing`.  The number of rows of data where dates were recorded but there are NA's for steps values is `r missingsteps`.  It is these rows with dates recorded that will be used going forward with mean for each interval replacing NA values.  The uniform spread of NA's across intervals should minimize the impact of replacement in this manner.  As can be seen in the histogram the new mean is `r smean2` which is nearly the same (less than .3% difference) and the median of `r smedian2` is unchanged.




```{r}

library("plyr")
library("lattice")
outdf$day <- weekdays(outdf$date)
outdf$wkend <- ifelse(outdf$day %in% c("Saturday", "Sunday"),"weekend", "weekday")
avgmeandf <- ddply(outdf, c("interval","wkend"), summarise, mean = mean(steps))
xyplot(mean ~ interval | wkend, avgmeandf, type = "l", layout = c(1,2), ylab = "Average Steps", xlab = "Interval")

```

Average steps per day for weekends vs weekdays over 5 minute intervals outlined above.
